\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{tikz}
\usetikzlibrary{graphs, quotes}

\title{HW1}
\author{Dev Goyal }
\date{February 2023}

\begin{document}


\maketitle

\section{Problem 1}

We're required to prove ($p \land (p \rightarrow q) \rightarrow q$) using resolution.

We will prove this by showing that the resolution of ($p \land (p \rightarrow q)) \land (\neg q$)[this is the negation of the expression] is unsatisfiable.

We will now convert the expression to CNF.

\begin{itemize}
    \item replace $p \rightarrow q $ with $\neg p \lor q$
    \item the expression is now $p \land (\neg p \lor q) \land \neg q$
\end{itemize}
we can now apply resolution to the expression. we resolve the last two terms tp get $p \land \neg p$ which is the empty clause thus the negation is unsatisfiable and our original expression is valid 

\section{Problem 2}
\begin{itemize}
    \item The CNF is $\{ \{ q, \neg r\}, \{ \neg p, r \},\{ \neg q, r, p\} \{p, q, \neg q \}, \{\neg r, q \}\}$
    \begin{itemize}
        \item resolving the first two terms we get: $\{ \{ q, \neg p\},\{ \neg q, r, p\}, \{p, q, \neg q \}, \{\neg r, q \} \} $
        \item resolving the third term we get: $\{ \{ q, \neg p\},\{ \neg q, r, p\}, \{\neg r, q \} \} $
        \item resolving first and second term we get: $\{ \{ \neg p, r, p\}, \{\neg r, q \} \} $
        \item resolving the first term: $\{ \{\neg r, q \} \} $
        
        This cannot be resolved further.
        
    \end{itemize}
    
    \item 
\end{itemize}

\section{ Problem 3}

the CNF in question 1 was $\{ \{ p\}, \{ \neg p, q \}, \{ \neg q\} \} $[ note this is the negation of the original expression]

% \begin{tikzpicture}
%     [sibling distance=10em,level distance=6em,
%     every node/.style={shape=rectangle,draw,align=center}]

%     \node{p}
%     child { node {q}};
% \end{tikzpicture}

We can represent this as a tree as follows following the order $p > \neg q > \neg p \lor q$:

\tikz \graph[grow down sep = 1cm, branch right=2cm, nodes={circle, draw}] { "p"  <- {
    "$\neg q $" [> "T"] <- { "$\neg p \lor q$" [> "T"], "$\neg q$" [> "F"]} ,
    "$ p $" [> "F"] <- {}
}};

using this tree we can derive the resolution proof tree as follows:

\tikz \graph[grow down sep = 1cm, branch right=2cm] { "$\{ \}$"  <- {
    "$ \{ \neg p \} $" [> "T"] <- { "$\{ \neg p, q \}$" [> "T"], "$ \{ \neg q \}$" [> "F"]} ,
    "$ \{ p \}$" [> "F"] <- {}
}};

As always, We start from the bottom children and keep resolving the terms until we reach the root. We can see that the root is empty and thus the expression is unsatisfiable.

thus our original expression is valid.




\end{document}